---
title: "Determining whether political affiliation and identifying as an environmentalist are independent"
author: "Sage Kime"
date: "12/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(here)
library(kableExtra)
library(broom)
```

### Overview

In this document, I explore and wrangle the conservation survey data set to test for independence between political affiliation and identification as an environmentalist. I will wrangle the data to only include the aforementioned variables. Then I will create a finalized table that shows the proportions and counts of the survey results we are examining. Finally, I will perform a chi square test to determine if a significant association exists, and I will state my overall takeaways from the data.

```{r, echo=FALSE}
# Read in data and keep only necessary variables

conservation_survey <- read.csv(here("data", "conservation_survey.csv")) %>% 
  select(ENV_ID, POL_OR) # use select function to pick which variables to keep
```



```{r, echo=FALSE}
# Add new column identifying political affiliation

party_subset <- conservation_survey %>% 
  mutate(POL_OR = case_when(
    between(POL_OR, 1, 3) ~ "Conservative",
    between(POL_OR, 5, 7) ~ "Liberal"
  )) %>% 
  na.omit()
```

### Survey responses

```{r, echo=FALSE}
# Create a finalized table w/ counts and proportions - add figure caption

survey_results <- party_subset %>% 
  janitor::tabyl(POL_OR, ENV_ID) %>% 
  rename("Yes" = "1", "No" = "2", "Unsure" = "3")

  survey_proportions <- survey_results %>% 
  adorn_percentages() %>% # use this to create a table with counts AND proportions
  adorn_pct_formatting(digits = 2) %>% # 
  adorn_ns()
  
  kbl(survey_proportions, col.names = c("", "Yes", "No", "Unsure"), caption = "Table 1. Survey responses show whether individuals identify as environmentalists, by political affiliation. Data source: Andrews Forest LTER Site and C. Batavia. 2019") %>% #rename to look tidy
    kable_styling(full_width = FALSE, "hover")
  
```



```{r, echo=FALSE}
# Make contingency table

survey_ct <- survey_results %>% 
  column_to_rownames(var = "POL_OR") #convert entries in a column to stored rownames
```

```{r, echo=FALSE}
# Run chi-square and tidy results

survey_x2 <- chisq.test(survey_ct) #perform chi square test

survey_tidy <- tidy(survey_x2) # use tidy function from broom package to make in-line referencing easier
```

### Summary

There is a significant association between political affiliation and identifying as an environmentalist ($\chi$^2^(`r survey_tidy$parameter`) = `r round(survey_tidy$statistic,2)`, *p* = `r round(survey_tidy$p.value, 3)`). Looking back at Table 1, we can see that 53.73% of liberals identified as environmentalists while only 33.65% of conservatives identified as environmentalists. Additionally, 50.24% of conservatives responded that they do NOT identify as environmentalists compared to only 24.47% of liberals. These differences are large enough to support our rejection of the null hypothesis and conclude that there is a significant association between the variables, with liberals being more likely to identify as environmentalists.

### Citation

Data source: Andrews Forest LTER Site and C. Batavia. 2019. Influences on charitable giving for conservation: Online survey data of 1,331 respondents across the US, August 2017 ver 3. Environmental Data Initiative. https://doi.org/10.6073/pasta/926e6270e324a1322a900da14d38b96c (Accessed 2020-12-06).
